---
title: "Elderly Fall Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Elderly Fall Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/laavanya/elderly-fall-prediction-and-detection?select=cStick.csv\>*

### Reference:

*\<Rachakonda, L. (2021). Elderly Fall Prediction and Detection Dataset. Retrieved from https://www.kaggle.com/datasets/laavanya/elderly-fall-prediction-and-detection?select=cStick.csv \>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load dataset
elderly_fall_data <- read.csv("elderly.csv", colClasses = c(
  Distance = "numeric",
  Pressure = "factor",
  HRV = "numeric",
  Sugar_level = "numeric",
  SpO2 = "numeric",
  Accelerometer = "factor",
  Decision = "factor"
))

# Display the structure of the dataset
str(elderly_fall_data)

# View the first few rows of the dataset
head(elderly_fall_data)

# View the dataset in a separate viewer window
View(elderly_fall_data)
```

## Measures of Frequency
```{r MOF}
# Calculate counts and percentages for Pressure levels
pressure_freq <- table(elderly_fall_data$Pressure)
pressure_percent <- prop.table(pressure_freq) * 100

# Calculate counts and percentages for Accelerometer readings
accelerometer_freq <- table(elderly_fall_data$Accelerometer)
accelerometer_percent <- prop.table(accelerometer_freq) * 100

# Calculate counts and percentages for Decision outcomes
decision_freq <- table(elderly_fall_data$Decision)
decision_percent <- prop.table(decision_freq) * 100

# Display the frequency tables
cat("Frequency of Pressure levels:\n")
print(pressure_freq)
cat("\nPercentage of Pressure levels:\n")
print(pressure_percent)

cat("\nFrequency of Accelerometer readings:\n")
print(accelerometer_freq)
cat("\nPercentage of Accelerometer readings:\n")
print(accelerometer_percent)

cat("\nFrequency of Decision outcomes:\n")
print(decision_freq)
cat("\nPercentage of Decision outcomes:\n")
print(decision_percent)
```

## Measures of Central Tendency
```{r MOCT}
# Calculate measures of central tendency for Distance
distance_mean <- mean(elderly_fall_data$Distance)
distance_median <- median(elderly_fall_data$Distance)
distance_mode <- as.numeric(names(sort(table(elderly_fall_data$Distance), decreasing = TRUE)[1]))

# Calculate measures of central tendency for HRV
hrv_mean <- mean(elderly_fall_data$HRV)
hrv_median <- median(elderly_fall_data$HRV)
hrv_mode <- as.numeric(names(sort(table(elderly_fall_data$HRV), decreasing = TRUE)[1]))

# Display the measures of central tendency
cat("Measures of Central Tendency for Distance:\n")
cat("Mean:", distance_mean, "\n")
cat("Median:", distance_median, "\n")
cat("Mode:", distance_mode, "\n\n")

cat("Measures of Central Tendency for HRV:\n")
cat("Mean:", hrv_mean, "\n")
cat("Median:", hrv_median, "\n")
cat("Mode:", hrv_mode, "\n\n")

```

## Measures of Distribution
```{r MOD}
# Calculate Measures of Distribution for Distance
distance_range <- range(elderly_fall_data$Distance)
distance_variance <- var(elderly_fall_data$Distance)
distance_sd <- sd(elderly_fall_data$Distance)
distance_quantiles <- quantile(elderly_fall_data$Distance, probs = c(0.25, 0.5, 0.75))

# Calculate Measures of Distribution for HRV
hrv_range <- range(elderly_fall_data$HRV)
hrv_variance <- var(elderly_fall_data$HRV)
hrv_sd <- sd(elderly_fall_data$HRV)
hrv_quantiles <- quantile(elderly_fall_data$HRV, probs = c(0.25, 0.5, 0.75))


# Display the Measures of Distribution
cat("Measures of Distribution for Distance:\n")
cat("Range:", distance_range, "\n")
cat("Variance:", distance_variance, "\n")
cat("Standard Deviation:", distance_sd, "\n")
cat("Quantiles (25th, 50th, 75th):", distance_quantiles, "\n\n")

cat("Measures of Distribution for HRV:\n")
cat("Range:", hrv_range, "\n")
cat("Variance:", hrv_variance, "\n")
cat("Standard Deviation:", hrv_sd, "\n")
cat("Quantiles (25th, 50th, 75th):", hrv_quantiles, "\n\n")

```

## Measures of Relationship
```{r MOR}
# Calculate correlation coefficients for numerical variables
correlation_matrix <- cor(elderly_fall_data[c("Distance", "HRV", "Sugar_level", "SpO2")])

# Display correlation matrix
cat("Correlation Matrix for Numerical Variables:\n")
print(correlation_matrix)

# Calculate contingency table for Pressure and Decision
pressure_decision_table <- table(elderly_fall_data$Pressure, elderly_fall_data$Decision)

# Display contingency table
cat("\nContingency Table for Pressure and Decision:\n")
print(pressure_decision_table)

# Calculate contingency table for Accelerometer and Decision
accelerometer_decision_table <- table(elderly_fall_data$Accelerometer, elderly_fall_data$Decision)

# Display contingency table
cat("\nContingency Table for Accelerometer and Decision:\n")
print(accelerometer_decision_table)
```

## ANOVA
```{r ANOVA}
# Load the required library for ANOVA
library(stats)

# Perform ANOVA for Distance across different levels of Pressure
anova_result <- aov(Distance ~ Pressure, data = elderly_fall_data)

# Summary of ANOVA results
summary(anova_result)
```

## Plots
```{r Plots }
# Load the required library for plotting
library(ggplot2)

# Create histograms for numerical variables
distance_plot <- ggplot(elderly_fall_data, aes(x = Distance)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Distance", x = "Distance", y = "Frequency")

hrv_plot <- ggplot(elderly_fall_data, aes(x = HRV)) +
  geom_histogram(binwidth = 5, fill = "lightgreen", color = "black") +
  labs(title = "Histogram of HRV", x = "HRV", y = "Frequency")

sugar_level_plot <- ggplot(elderly_fall_data, aes(x = Sugar_level)) +
  geom_histogram(binwidth = 5, fill = "salmon", color = "black") +
  labs(title = "Histogram of Sugar Level", x = "Sugar Level", y = "Frequency")

spo2_plot <- ggplot(elderly_fall_data, aes(x = SpO2)) +
  geom_histogram(binwidth = 5, fill = "lightyellow", color = "black") +
  labs(title = "Histogram of SpO2", x = "SpO2", y = "Frequency")

# Load the required library for arranging plots
library(gridExtra)

# Combine plots into a single display
grid.arrange(distance_plot, hrv_plot, sugar_level_plot, spo2_plot, ncol = 2)

# Create box plot for Distance across different Pressure levels
distance_pressure_plot <- ggplot(elderly_fall_data, aes(x = Pressure, y = Distance, fill = Pressure)) +
  geom_boxplot() +
  labs(title = "Box Plot of Distance by Pressure Levels", x = "Pressure", y = "Distance")

# Create box plot for HRV across different Pressure levels
hrv_pressure_plot <- ggplot(elderly_fall_data, aes(x = Pressure, y = HRV, fill = Pressure)) +
  geom_boxplot() +
  labs(title = "Box Plot of HRV by Pressure Levels", x = "Pressure", y = "HRV")

# Create violin plot for Sugar Level across different Pressure levels
sugar_pressure_plot <- ggplot(elderly_fall_data, aes(x = Pressure, y = Sugar_level, fill = Pressure)) +
  geom_violin() +
  labs(title = "Violin Plot of Sugar Level by Pressure Levels", x = "Pressure", y = "Sugar Level")

# Create violin plot for SpO2 across different Pressure levels
spo2_pressure_plot <- ggplot(elderly_fall_data, aes(x = Pressure, y = SpO2, fill = Pressure)) +
  geom_violin() +
  labs(title = "Violin Plot of SpO2 by Pressure Levels", x = "Pressure", y = "SpO2")

# Combine plots into a single display
grid.arrange(distance_pressure_plot, hrv_pressure_plot, sugar_pressure_plot, spo2_pressure_plot, ncol = 2)


```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Count missing values in each column
missing_values <- colSums(is.na(elderly_fall_data))

# Display columns with missing values
print("Columns with Missing Values:")
print(names(missing_values[missing_values > 0]))

# Display total number of missing values
print("Total Number of Missing Values:")
print(sum(missing_values))
```

## Transformation
```{r Transformation}
# Apply logarithmic transformation to Sugar_level variable
elderly_fall_data$Sugar_level_log <- log(elderly_fall_data$Sugar_level)

# Plot the histogram of the original and transformed Sugar_level variable
ggplot(elderly_fall_data, aes(x = Sugar_level)) +
  geom_histogram(binwidth = 5, fill = "salmon", color = "black") +
  labs(title = "Histogram of Original Sugar Level", x = "Sugar Level", y = "Frequency")

ggplot(elderly_fall_data, aes(x = Sugar_level_log)) +
  geom_histogram(binwidth = 0.1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Transformed Sugar Level (Log)", x = "Log(Sugar Level)", y = "Frequency")

```

# Training Models
## Data splitting
```{r Data splitting}
# Load the required library for data splitting
library(caret)

# Set seed for reproducibility
set.seed(123)

# Specify the percentage of data to be allocated for testing (e.g., 80% training, 20% testing)
split_percentage <- 0.8

# Split the dataset into training and testing sets
training_index <- createDataPartition(elderly_fall_data$Decision, p = split_percentage, list = FALSE)
training_data <- elderly_fall_data[training_index, ]
testing_data <- elderly_fall_data[-training_index, ]

# Print the dimensions of the training and testing sets
print("Dimensions of the training set:")
print(dim(training_data))
print("Dimensions of the testing set:")
print(dim(testing_data))
```

## Bootstrapping
```{r Bootstrapping}
# Define the number of bootstrap samples
num_bootstraps <- 1000

# Create an empty vector to store bootstrap statistics
bootstrap_statistics <- numeric(num_bootstraps)

# Perform bootstrapping
for (i in 1:num_bootstraps) {
  # Resample the dataset with replacement
  bootstrap_sample <- elderly_fall_data[sample(nrow(elderly_fall_data), replace = TRUE), ]
  
  # Calculate the statistic of interest (e.g., mean, median, etc.)
  # For example, let's calculate the mean of the Distance variable
  bootstrap_statistics[i] <- mean(bootstrap_sample$Distance)
}

# Compute the confidence interval
confidence_interval <- quantile(bootstrap_statistics, c(0.025, 0.975))

# Print the confidence interval
print(confidence_interval)
```

## RF Model
```{r RF Model}
# Load required libraries
library(caret)
library(randomForest)

# Define train control using cross-validation
train_control <- trainControl(method = "cv", number = 10)  # 10-fold cross-validation

# Train a Random Forest Regression model
rf_model <- train(Decision ~ ., data = elderly_fall_data, method = "rf", trControl = train_control)

# Print the trained Random Forest model
print(rf_model)
```

## GBM Model
```{r GBM Model}
# Load the required library for modeling
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define the training control
train_control <- trainControl(method = "cv", number = 10)

# Train a Gradient Boosting Regression model
gbm_model <- train(Decision ~ ., data = elderly_fall_data, method = "gbm", trControl = train_control)

# Print the trained Gradient Boosting model
print(gbm_model)
```

## SVR Model
```{r SVR Model}
# Load the required library for modeling
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define the training control
train_control <- trainControl(method = "cv", number = 10)

# Train a Support Vector Regression (SVR) model
svr_model <- train(Decision ~ ., data = elderly_fall_data, method = "svmRadial", trControl = train_control)

# Print the trained SVR model
print(svr_model)
```

## Performance Comparison
```{r Performance Comparison}
# Load the required library for modeling
library(caret)

# Set seed for reproducibility
set.seed(123)

# Define the training control
train_control <- trainControl(method = "cv", number = 10)

# Define the models
models <- list(
  SVM = train(Decision ~ ., data = elderly_fall_data, method = "svmRadial", trControl = train_control),
  RF = train(Decision ~ ., data = elderly_fall_data, method = "rf", trControl = train_control),
  GBM = train(Decision ~ ., data = elderly_fall_data, method = "gbm", trControl = train_control)
)

# Compare model performance using resamples
resamples_list <- resamples(models)

# Summarize and compare the models
summary_resamples <- summary(resamples_list)
print(summary_resamples)
```

## Saving Model
```{r Saving Model}
# Load the saved model
loaded_elderly_rf_model <- readRDS("./models/elderly_rf_model.rds")

# Prepare new data for prediction
new_elderly_data <- data.frame(
  Distance = 25.54,
  Pressure = factor(1, levels = c(0, 1, 2)),  # Assuming factor levels are 0, 1, and 2
  HRV = 101.396,
  Sugar_level = 61.08,
  SpO2 = 87.77,
  Accelerometer = factor(1, levels = c(0, 1)),  # Assuming factor levels are 0 and 1
  Decision = factor(0, levels = c(0, 1, 2))  # Assuming factor levels are 0, 1, and 2
)

# Use the loaded model to make predictions for elderly data
predictions_elderly_loaded_model <- predict(loaded_elderly_rf_model, newdata = new_elderly_data)

# Print predictions
print(predictions_elderly_loaded_model)

```

